---
title: "Exploratory Data Analysis"
output:
  html_document:
    df_print: paged
---

##Exploratory Data Analysis- Section 4

###Analytical Graphics
Principle 1: show comparisons, and don't forget your control
Principle 2: Show causality, mechanism, explanation for differences
Principle 3: show multi-variate data
Principle 4: Show what you need, and don't be limited by the tools
Principle 5: describe and document your evidence (code, labels, etc...)
Principle 6: content is king- the story you tell is most important

##Exploratory Graphs
Why? Look at trends, debug an analysis, communicate results, suggest modelling strategies.
    - made quickly, many are made at once, axes are cleaned later, and colors will change. 
    
```{r}
## par() changes the global graphics parameters for all graphics. Requires a reset by using the code below: 
old.par <- par(mfrow = c(1,1), mfcol = c(1,1), mar = c(5, 4, 4, 2))
par(old.par) ##This code allows you to reset the parameters of a plot in R. 

## The list below are all par() commands. 
# ls() is the orientation of axis labels on a plot
# bg() is the background color
# mar() defines the margin size
# oma() is the outer margin size
# mfrow() is numbers of plots per row, column, filled row-wise. 
# mfcol() is numbers of plots per row. column, filled column-wise. 

boxplot(mtcars$mpg, col= "blue") ##example of a boxplot
abline(h = max(mtcars$mpg), col= "orange", lwd = 2)  ##h here is horizontal line addition
abline(h = mean(mtcars$mpg), col ="green", lwd = 2)  ## lwd is the line width/thickness

##_______________________
boxplot(mpg ~ cyl, data = mtcars, col = "orange")  ##plotting the relationship " ~ " between to variables, where mpy is y axis, and cyl is x axis. 
abline(h = mean(mtcars$mpg), col = "blue", lwd = 2)

##_______________________
mtcars2 <- transform(mtcars, cyl= factor(cyl))
boxplot(disp ~ cyl, data = mtcars2, xlab = "Cylinders", ylab = "Displacement") ##labeling axes in base R. 

##_______________________
hist(mtcars$mpg, col= "orange", breaks = 20, main = "Miles per Gallon accross cars") ##example of a histogram
rug(mtcars$mpg)
abline(v = mean(mtcars$mpg), col ="purple", lwd = 3)  ##v stands for vertical line. 
abline(v = max(mtcars$mpg), col = "yellow", lwd = 2)

##_______________________
barplot(table(mtcars$gear), col= "turquoise", main = "Number of Gears per Car")

##_______________________
mtcars <- na.omit(mtcars)
with(mtcars, plot(mpg, disp, col=mpg))
with(subset(mtcars, disp > 400), points(mpg, disp, col = "black")) ##xy scatter plot comparing two column variables. we are taking a subset of the mtcars where displacement is greater than 400, and making those points black.  

##_______________________
##Adding more details to standard plot, e.g. using the legend expression, and adding regression. 
with(mtcars, plot(mpg, disp, main = "Association Between MPG and Displacement", type = "n"))
with(subset(mtcars, cyl == 4), points(mpg, disp, col = "green"))
with(subset(mtcars, cyl != 4), points(mpg, disp, col = "red"))
legend("topright", pch = 1, col = c("green", "red"), legend = c("4 Cylinders", "More than 4 cylinders")) #pch changes the characteristic of the circle/point in the plot. 
model <- lm(disp ~ mpg, mtcars) ##creating a liner model between two varialbes, mpg and displacement
abline(model, col = "black", lty = 2)


#abline(h = mean(mtcars$weight), col="orange", lty = 2)  ##lty is a dotted line. 
#abline(v = mean(mtcars$mpg), col="yellow", lty=2)

##_______________________
par(mfrow = c(3,1), mar = c(5,4,2,1))  ##more to learn here  ##starting at bottom and moving clockwise is how margins are oriented. For mfrow, it states one row, three column.
mtcars$gear <- as.integer(mtcars$gear)
with(subset(mtcars, gear == 3), plot(mpg, wt, col = wt, main = "Three Gears"))
with(subset(mtcars, gear == 4), plot(mpg, wt, col = wt, main = "Four Gears"))
with(subset(mtcars, gear == 5), plot(mpg, wt, col = wt, main = "Five Gears"))

##_______________________
old.par <- par(mfrow = c(1,1), mfcol = c(1,1), mar = c(5, 4, 4, 2))
par(old.par) ##resetting
par(mfrow = c(1,3), oma = c(0,0,2,0))  ##gives us a little room on the outer margin for a main title/header.
with(airquality, {
  plot(Wind, Ozone, main = "Ozone vs. Wind")
  plot(Solar.R, Ozone, main = "Ozone vs. Solar Radiation")
  plot(Temp, Ozone, main = "Ozone and Temperature")
  mtext("Ozone vs. Weather in NYC", outer = TRUE) ##adds a main title/header to this series of three plots. 
})
##_______________________

##Standard plot systems
old.par <- par(mfrow = c(1,1), mfcol = c(1,1), mar = c(5, 4, 4, 2))
par(old.par) ##This code allows you to reset the parameters of a plot in R. 
with(cars, plot(speed, dist))

##_______________________
##Standard lattice system
#e.g. xyplot and bwplot
library(lattice)
xyplot(mpg ~ disp | gear, data = mtcars, layout = c(4,1)) ##separating x/y plot by gear

##_______________________
##ggplot2 plotting
library(ggplot2)
data(mtcars)
qplot(disp, mpg, data = mtcars)

```

### Exporting/Copying plots to files
Once the plot has been generated...
dev.copy(png, file = "filename.png")
dev.off()

### Histogram Example, Power Grid

```{r}
##code for histogram of power data, UC Irvine Machine Learning Repository
data_table <- read.table("./data/household_power_consumption.txt", header = TRUE, sep = ";", na.strings = "?")  ##separated by semi colons, and the na's are marked as "?".
data_table$Date_Time <- paste(data_table$Date, data_table$Time)  ##Paste the first two rows together and renames them as Date_Time
      data_table$Global_active_power <- as.numeric(data_table$Global_active_power)  ##changes needed data to numeric. 
data_table$Date_Time <- strptime(data_table$Date_Time, "%d/%m/%Y %H:%M:%S")  #defining time string as time. 
         data_table <- data_table[data_table$Date %in% c("1/2/2007", "2/2/2007"), ]  ##subsets the data for the two dates in question. 
               head(data_table, n=20) ##quick gut check

hist(data_table$Global_active_power, col= "red", breaks = 14, main = "Global Active Power"
      , xlab = "Global Active Power (kilowatts)")
            dev.copy(png, file = "plot1_R.png")
            dev.off()
```
### Plotting example, Line Plot, Power Grid

```{r}
##code for line plot of power data, UC Irvine Machine Learning Repository
data_table <- read.table("./data/household_power_consumption.txt", header = TRUE, sep = ";", na.strings = "?")  ##separated by semi colons, and the na's are marked as "?".
data_table$Date_Time <- paste(data_table$Date, data_table$Time)  ##Paste the first two rows together and renames them as Date_Time
      data_table$Global_active_power <- as.numeric(data_table$Global_active_power)  ##changes needed data to numeric. 
data_table$Date_Time <- strptime(data_table$Date_Time, "%d/%m/%Y %H:%M:%S")  #defining time string as time. 
       data_table <- data_table[data_table$Date %in% c("1/2/2007", "2/2/2007"), ]  ##subsets the data for the two dates in question. 
          head(data_table, n=20) ##quick gut check
          
with(data_table, plot(Date_Time, Global_active_power, xlab = "", ylab = "Global Active Power (kilowatts)", type ="l", lwd = 2)) ##defining the plot type as line or "l", and width of 2. 
      dev.copy(png, file = "plot2_R.png")
          dev.off()
```
###Plotting + Lines (multi-column) Example, Power Grid

```{r}
##code for three-variable plot of metering data, UC Irvine Machine Learning Repository
data_table <- read.table("./data/household_power_consumption.txt", header = TRUE, sep = ";", na.strings = "?")  ##separated by semi colons, and the na's are marked as "?".
    data_table$Date_Time <- paste(data_table$Date, data_table$Time)  ##Paste the first two rows together and renames them as Date_Time
data_table$Global_active_power <- as.numeric(data_table$Global_active_power)  ##changes needed data to numeric. 
    data_table$Date_Time <- strptime(data_table$Date_Time, "%d/%m/%Y %H:%M:%S")  #defining time string as time. 
        data_table <- data_table[data_table$Date %in% c("1/2/2007", "2/2/2007"), ]  ##subsets the data for the two dates in question. 
            head(data_table, n=20) ##quick gut check
            
with(data_table, plot(Date_Time, Sub_metering_1, col= "black", xlab = "", ylab ="Energy sub metering", type = "l"))
                      lines(data_table$Date_Time, data_table$Sub_metering_2, col="orange", type="l")
                      lines(data_table$Date_Time, data_table$Sub_metering_3, col="blue", type="l")
                          legend("topright", c("Sub_metering_1", "Sub_metering_2", "Sub_metering_3"), col = c("black", "orange", "blue"), lty = 1)
                              dev.copy(png, file = "plot3_R.png")
                               dev.off() 
```

###2x2 Plot Example, Power Grid

```{r}
##code for 2x2 plot of metering data, UC Irvine Machine Learning Repository
data_table <- read.table("./data/household_power_consumption.txt", header = TRUE, sep = ";", na.strings = "?")  ##separated by semi colons, and the na's are marked as "?".
    data_table$Date_Time <- paste(data_table$Date, data_table$Time)  ##Paste the first two rows together and renames them as Date_Time
       data_table$Global_active_power <- as.numeric(data_table$Global_active_power)  ##changes needed data to numeric. 
       data_table$Global_reactive_power <- as.numeric(data_table$Global_reactive_power)
       data_table$Voltage <- as.numeric(data_table$Voltage)
data_table$Date_Time <- strptime(data_table$Date_Time, "%d/%m/%Y %H:%M:%S")  #defining time string as time. 
      data_table <- data_table[data_table$Date %in% c("1/2/2007", "2/2/2007"), ]  ##subsets the data for the two dates in question. 
          head(data_table, n=20) ##quick gut check
          

par(mfrow = c(2,2))
with(data_table, plot(Date_Time, Global_active_power, xlab = "", ylab = "Global Active Power (kilowatts)", type ="l", lwd = 2))
        with(data_table, plot(Date_Time, Voltage, xlab = "", ylab = "Voltage", type ="l", lwd = 1))
            with(data_table, plot(Date_Time, Sub_metering_1, col= "black", xlab = "", ylab ="Energy sub metering", type = "l"))
            lines(data_table$Date_Time, data_table$Sub_metering_2, col="orange", type="l")
            lines(data_table$Date_Time, data_table$Sub_metering_3, col="blue", type="l")
            legend("topright", c("Sub_metering_1", "Sub_metering_2", "Sub_metering_3"), col = c("black", "orange", "blue"))
                  with(data_table, plot(Date_Time, Global_reactive_power, xlab = "", type ="l", lwd = 1))
                      dev.copy(png, file = "plot4_R.png")
                       dev.off() 
```

