---
title: "Exploratory Data Analysis"
output:
  html_document:
    df_print: paged
---

##Exploratory Data Analysis- Section 4

###Analytical Graphics
Principle 1: show comparisons, and don't forget your control
Principle 2: Show causality, mechanism, explanation for differences
Principle 3: show multi-variate data
Principle 4: Show what you need, and don't be limited by the tools
Principle 5: describe and document your evidence (code, labels, etc...)
Principle 6: content is king- the story you tell is most important

##Exploratory Graphs
Why? Look at trends, debug an analysis, communicate results, suggest modelling strategies.
    - made quickly, many are made at once, axes are cleaned later, and colors will change. 
    
```{r}
## par() changes the global graphics parameters for all graphics. Requires a reset by using the code below: 
old.par <- par(mfrow = c(1,1), mfcol = c(1,1), mar = c(5, 4, 4, 2))
par(old.par) ##This code allows you to reset the parameters of a plot in R. 

## The list below are all par() commands. 
# ls() is the orientation of axis labels on a plot
# bg() is the background color
# mar() defines the margin size
# oma() is the outer margin size
# mfrow() is numbers of plots per row, column, filled row-wise. 
# mfcol() is numbers of plots per row. column, filled column-wise. 

boxplot(mtcars$mpg, col= "blue") ##example of a boxplot
abline(h = max(mtcars$mpg), col= "orange", lwd = 2)  ##h here is horizontal line addition
abline(h = mean(mtcars$mpg), col ="green", lwd = 2)  ## lwd is the line width/thickness

##_______________________
boxplot(mpg ~ cyl, data = mtcars, col = "orange")  ##plotting the relationship " ~ " between to variables, where mpy is y axis, and cyl is x axis. 
abline(h = mean(mtcars$mpg), col = "blue", lwd = 2)

##_______________________
mtcars2 <- transform(mtcars, cyl= factor(cyl))
boxplot(disp ~ cyl, data = mtcars2, xlab = "Cylinders", ylab = "Displacement") ##labeling axes in base R. 

##_______________________
hist(mtcars$mpg, col= "orange", breaks = 20, main = "Miles per Gallon accross cars") ##example of a histogram
rug(mtcars$mpg)
abline(v = mean(mtcars$mpg), col ="purple", lwd = 3)  ##v stands for vertical line. 
abline(v = max(mtcars$mpg), col = "yellow", lwd = 2)

##_______________________
barplot(table(mtcars$gear), col= "turquoise", main = "Number of Gears per Car")

##_______________________
mtcars <- na.omit(mtcars)
with(mtcars, plot(mpg, disp, col=mpg))
with(subset(mtcars, disp > 400), points(mpg, disp, col = "black")) ##xy scatter plot comparing two column variables. we are taking a subset of the mtcars where displacement is greater than 400, and making those points black.  

##_______________________
##Adding more details to standard plot, e.g. using the legend expression, and adding regression. 
with(mtcars, plot(mpg, disp, main = "Association Between MPG and Displacement", type = "n"))
with(subset(mtcars, cyl == 4), points(mpg, disp, col = "green"))
with(subset(mtcars, cyl != 4), points(mpg, disp, col = "red"))
legend("topright", pch = 1, col = c("green", "red"), legend = c("4 Cylinders", "More than 4 cylinders")) #pch changes the characteristic of the circle/point in the plot. 
model <- lm(disp ~ mpg, mtcars) ##creating a liner model between two varialbes, mpg and displacement
abline(model, col = "black", lty = 2)


#abline(h = mean(mtcars$weight), col="orange", lty = 2)  ##lty is a dotted line. 
#abline(v = mean(mtcars$mpg), col="yellow", lty=2)

##_______________________
par(mfrow = c(3,1), mar = c(5,4,2,1))  ##more to learn here  ##starting at bottom and moving clockwise is how margins are oriented. For mfrow, it states one row, three column.
mtcars$gear <- as.integer(mtcars$gear)
with(subset(mtcars, gear == 3), plot(mpg, wt, col = wt, main = "Three Gears"))
with(subset(mtcars, gear == 4), plot(mpg, wt, col = wt, main = "Four Gears"))
with(subset(mtcars, gear == 5), plot(mpg, wt, col = wt, main = "Five Gears"))

##_______________________
old.par <- par(mfrow = c(1,1), mfcol = c(1,1), mar = c(5, 4, 4, 2))
par(old.par) ##resetting
par(mfrow = c(1,3), oma = c(0,0,2,0))  ##gives us a little room on the outer margin for a main title/header.
with(airquality, {
  plot(Wind, Ozone, main = "Ozone vs. Wind")
  plot(Solar.R, Ozone, main = "Ozone vs. Solar Radiation")
  plot(Temp, Ozone, main = "Ozone and Temperature")
  mtext("Ozone vs. Weather in NYC", outer = TRUE) ##adds a main title/header to this series of three plots. 
})
##_______________________

##Standard plot systems
old.par <- par(mfrow = c(1,1), mfcol = c(1,1), mar = c(5, 4, 4, 2))
par(old.par) ##This code allows you to reset the parameters of a plot in R. 
with(cars, plot(speed, dist))

##_______________________
##Standard lattice system
#e.g. xyplot and bwplot
library(lattice)
xyplot(mpg ~ disp | gear, data = mtcars, layout = c(4,1)) ##separating x/y plot by gear

##_______________________
##ggplot2 plotting
library(ggplot2)
data(mtcars)
qplot(disp, mpg, data = mtcars)

```

### Exporting/Copying plots to files
Once the plot has been generated...
dev.copy(png, file = "filename.png")
dev.off()

### Histogram Example, Power Grid

```{r}
##code for histogram of power data, UC Irvine Machine Learning Repository
data_table <- read.table("./data/household_power_consumption.txt", header = TRUE, sep = ";", na.strings = "?")  ##separated by semi colons, and the na's are marked as "?".
data_table$Date_Time <- paste(data_table$Date, data_table$Time)  ##Paste the first two rows together and renames them as Date_Time
      data_table$Global_active_power <- as.numeric(data_table$Global_active_power)  ##changes needed data to numeric. 
data_table$Date_Time <- strptime(data_table$Date_Time, "%d/%m/%Y %H:%M:%S")  #defining time string as time. 
         data_table <- data_table[data_table$Date %in% c("1/2/2007", "2/2/2007"), ]  ##subsets the data for the two dates in question. 
               head(data_table, n=20) ##quick gut check

hist(data_table$Global_active_power, col= "red", breaks = 14, main = "Global Active Power"
      , xlab = "Global Active Power (kilowatts)")
            dev.copy(png, file = "plot1_R.png")
            dev.off()
```
### Plotting example, Line Plot, Power Grid

```{r}
##code for line plot of power data, UC Irvine Machine Learning Repository
data_table <- read.table("./data/household_power_consumption.txt", header = TRUE, sep = ";", na.strings = "?")  ##separated by semi colons, and the na's are marked as "?".
data_table$Date_Time <- paste(data_table$Date, data_table$Time)  ##Paste the first two rows together and renames them as Date_Time
      data_table$Global_active_power <- as.numeric(data_table$Global_active_power)  ##changes needed data to numeric. 
data_table$Date_Time <- strptime(data_table$Date_Time, "%d/%m/%Y %H:%M:%S")  #defining time string as time. 
       data_table <- data_table[data_table$Date %in% c("1/2/2007", "2/2/2007"), ]  ##subsets the data for the two dates in question. 
          head(data_table, n=20) ##quick gut check
          
with(data_table, plot(Date_Time, Global_active_power, xlab = "", ylab = "Global Active Power (kilowatts)", type ="l", lwd = 2)) ##defining the plot type as line or "l", and width of 2. 
      dev.copy(png, file = "plot2_R.png")
          dev.off()
```
###Plotting + Lines (multi-column) Example, Power Grid

```{r}
##code for three-variable plot of metering data, UC Irvine Machine Learning Repository
data_table <- read.table("./data/household_power_consumption.txt", header = TRUE, sep = ";", na.strings = "?")  ##separated by semi colons, and the na's are marked as "?".
    data_table$Date_Time <- paste(data_table$Date, data_table$Time)  ##Paste the first two rows together and renames them as Date_Time
data_table$Global_active_power <- as.numeric(data_table$Global_active_power)  ##changes needed data to numeric. 
    data_table$Date_Time <- strptime(data_table$Date_Time, "%d/%m/%Y %H:%M:%S")  #defining time string as time. 
        data_table <- data_table[data_table$Date %in% c("1/2/2007", "2/2/2007"), ]  ##subsets the data for the two dates in question. 
            head(data_table, n=20) ##quick gut check
            
with(data_table, plot(Date_Time, Sub_metering_1, col= "black", xlab = "", ylab ="Energy sub metering", type = "l"))
                      lines(data_table$Date_Time, data_table$Sub_metering_2, col="orange", type="l")
                      lines(data_table$Date_Time, data_table$Sub_metering_3, col="blue", type="l")
                          legend("topright", c("Sub_metering_1", "Sub_metering_2", "Sub_metering_3"), col = c("black", "orange", "blue"), lty = 1)
                              dev.copy(png, file = "plot3_R.png")
                               dev.off() 
```

###2x2 Plot Example, Power Grid

```{r}
##code for 2x2 plot of metering data, UC Irvine Machine Learning Repository
data_table <- read.table("./data/household_power_consumption.txt", header = TRUE, sep = ";", na.strings = "?")  ##separated by semi colons, and the na's are marked as "?".
    data_table$Date_Time <- paste(data_table$Date, data_table$Time)  ##Paste the first two rows together and renames them as Date_Time
       data_table$Global_active_power <- as.numeric(data_table$Global_active_power)  ##changes needed data to numeric. 
       data_table$Global_reactive_power <- as.numeric(data_table$Global_reactive_power)
       data_table$Voltage <- as.numeric(data_table$Voltage)
data_table$Date_Time <- strptime(data_table$Date_Time, "%d/%m/%Y %H:%M:%S")  #defining time string as time. 
      data_table <- data_table[data_table$Date %in% c("1/2/2007", "2/2/2007"), ]  ##subsets the data for the two dates in question. 
          head(data_table, n=20) ##quick gut check
          

par(mfrow = c(2,2))
with(data_table, plot(Date_Time, Global_active_power, xlab = "", ylab = "Global Active Power (kilowatts)", type ="l", lwd = 2))
        with(data_table, plot(Date_Time, Voltage, xlab = "", ylab = "Voltage", type ="l", lwd = 1))
            with(data_table, plot(Date_Time, Sub_metering_1, col= "black", xlab = "", ylab ="Energy sub metering", type = "l"))
            lines(data_table$Date_Time, data_table$Sub_metering_2, col="orange", type="l")
            lines(data_table$Date_Time, data_table$Sub_metering_3, col="blue", type="l")
            legend("topright", c("Sub_metering_1", "Sub_metering_2", "Sub_metering_3"), col = c("black", "orange", "blue"))
                  with(data_table, plot(Date_Time, Global_reactive_power, xlab = "", type ="l", lwd = 1))
                      dev.copy(png, file = "plot4_R.png")
                       dev.off() 
```

##Lattice Plotting
xyplot( y~ x | f * g, data) is the standard notation for a lattice plot. 
left of ~ is trhe y-axis, and on the right is the x-axis
f and g are conditioning variables where the * indicates an interaction between two variables
```{r}
library(lattice)
xyplot(mpg ~ disp, data =mtcars) ##simple scatter plot example using mtcars. 
xyplot(Ozone ~ Wind, data = airquality)

##_____________________
airquality <- transform(airquality, Month = factor(Month))  ##Month was turned into a factor that we will ultimately parse the data into. 
xyplot(Ozone ~ Wind | Month, data = airquality, layout = c(5,1))  ##xyplot, where month was listed as a conditioning variable. 

```

##ggplot2 plotting
Grammar of graphics by Leland Wilkinson (ggplot2.org)
aesthetics: size, shape, color
geoms: points, lines
###qplot
```{r}
library(ggplot2) 
str(mpg) ##example dataset that comes with ggplot2
qplot(displ, hwy, data = mpg)  ##most basic of "base" plots for ggplot2. displ = x-coordinate, hwy = y-coordinate. 
qplot(displ, hwy, data = mpg, color = drv)  ##adds a legend where colors are specific to variable drv (front, rear or 4w)
qplot(displ, hwy, data = mpg, geom = c("point", "smooth"))  ## grey = 95% confidence interval, and "smooth" is simply the function for the line. 
qplot( hwy, data = mpg, fill = drv)  ##when only one variable is defined, it will produce a histogram. fill is an alternative to color when using a histogpram. 

##Facets_______________
qplot(displ, hwy, data = mpg, facets = .~drv) ## here, the facets function splits "." or everything by ~ the variable drv
qplot(hwy, data= mpg, facets = drv~., binwidth= 2, color = drv)## here, histogram where facet used to separate drv ~ by "." or everything

##Geoms______________________
qplot(hwy, data = mpg, geom="density", color = drv) ##example of geom "density" which is a smooth plot alternative to histogram, can accept y variable only, notx and y

qplot(displ, hwy, data = mpg, color = drv) + geom_smooth(method = "lm")
qplot(displ, hwy, data = mpg, facets = .~drv, color = drv) + geom_smooth(method = "lm") ## using facets to separate out the regression by drv. lm for linear model.
qplot(displ, hwy, data = mpg, facets = .~ cyl, color = cyl, geom = c("point", "smooth"), method = "lm")  ##again, point and smooth gives us the confidence interval, along with a line "lm" for linear model. 

##______________________
```

### ggplot

```{r}
g <- ggplot(mpg, aes(hwy, cyl))  ##aes stands for the aesthetic elements, i.e. what and it's orientation.
summary(g)
##print(g) will return an error because there is  insufficient information to provide an output. 
p <- g + geom_point()  ##point layer added to g
print(p)
 g + geom_point() + geom_smooth()  ##smooths out and adds confidence bands. 
 g + geom_point() + geom_smooth(method = "lm") ##adds a linear model best-fit line
 
 ##___________________
 h <- ggplot(mpg, aes(displ, drv))
 h + geom_point(color = "forestgreen", size = 4, alpha= 1/2) + ##color of this variable is a constant - green
   #facet_grid(. ~ cyl) + 
   geom_smooth(method = "lm") + 
   xlab("Displacement") + ##labels the x axis
  ylab("Drive_Type") +
   ggtitle("Engine Attributes")
 
        h <- ggplot(mpg, aes(displ, cty))
        h + geom_point(aes(color = cyl), size = 4, alpha= 1/2) +   ###assigning a color to a variable requires aes(color = column name), and the alpha is the transparency, size is just that. 
          theme_classic(base_family = "Avenir", base_size = 14) + ##changes the base font for the entire plot
           geom_smooth(size = 1.5, linetype = 6, color = "orange", method = "lm") + 
           xlab("Displacement") + ##labels the x axis
            ylab("City_MPG") +
            ggtitle("Engine Attributes") +
            ylim(-5, 30)  ##sets a limit to the y axis, but in ggplot, that will subset the data and remove outlier points. 
                ## + coor_cartesian(ylim = c(-5, 30))  This is a way of preventing sub-setting and outliers are included in the finished plot. 
  
        ##facet_wrap( column/category ~ column/category2, nrow = y, ncol = x)
```

##Heirarchical Clustering [i.e. cluster analysis]
- How do we define close?
    - Continuous = euclidian distance (straight line) a^2 + b^2 = c^2 
    - Continuous = correlation similarity
    - Binary = manhattan distance. |A1 - A2| + |B1 - B2| + .... |Z1 - Z2| absolute values - think walking city blocks.
- How do we group things? 
- How to we visualize the groupings? 
- How do we interpret the grouping? 

Taking an agglomerative approach (bottom up), find closes two things, put them into larger clusters, find next closest, etc...
    - Required a defined distance
    - A merging approach
  Produced: a tree showing how close things are to each other. 

e.g. dist(mpg) gives you the pairwise differences between each variable in each column. 

###Dendrogram - rudimentary
```{r}
distxy <- dist(mpg$cty)
hClustering <- hclust(distxy)
plot(hClustering)  ##this clustering does not necessarily demonsrate anything significant in terms of output or trends, but will at least produce a figure. 
          #clusterMatrix <- as.matrix(hClustering[ , 8:9])
          #heatmap(clusterMatrix)



set.seed(1234)
par(mar = c(0,0,0,0))
x <- rnorm(12, mean= rep(1:3, each = 4), sd = 0.2)
y <- rnorm(12, mean = rep(c(1,2,1), each = 4), sd = 0.2)
plot(x,y, col= "orange", pch= 5, cex = 2)
text(x + 0.08, y + 0.08, labels = as.character(1:12))
      Data_thing <- data.frame( x=x, y=y)
      data_dist <- dist(Data_thing)
      thing_cluster <- hclust(data_dist)
      plot(thing_cluster)
           thing_matrix <- as.matrix(Data_thing) [sample(1:12), ]
           heatmap(thing_matrix)  

```

##k-means Clustering & Dimension reduction
Partitions
  -fix a number of clusters
  - get "centroids" of each cluster
  - assign things to closest centroid
  - recalculate centroids and iterate
Requires
  - A defined distance metric
  - a number of clusters
  - an initial guess as to cluster centroids
Produces
  - Final estimate of cluster and assignments
  
###kmeans Clustering  
```{r}
##Generating random data here
set.seed(1234)
par(mar = c(0,0,0,0))
x <- rnorm(12, mean= rep(1:3, each = 4), sd = 0.2)
y <- rnorm(12, mean = rep(c(1,2,1), each = 4), sd = 0.2)
plot(x,y, col= "orange", pch= 5, cex = 2)
text(x + 0.08, y + 0.08, labels = as.character(1:12))

##kmeans function/argument for clustering analyses
data_thing <- data.frame(x = x, y = y)
kmeansObject <- kmeans(data_thing, centers = 3)  ##caling kmeans argument, three centers/groups
names(kmeansObject)
par(mar = rep(0.2, 4))  #adjustes the margins
plot( x, y, col= kmeansObject$cluster, pch = 19, cex = 2)  #plots the kmeans function, calling out the clusters
points(kmeansObject$centers, col = 1:3, pch = 3, lwd = 3)  # plots the points that represent the cluster centers


```
 
### Principal components analysis and singular decomposition 
Not much to report here, lol. 

### Working with color in R
```{r}
library(grDevices) ##takes blending of primary and other colors and blend them together. 
# colorRamp  (e.g. gray)
# colorRampPalette  (e.g. heat.colors)
# colors()
color1 <- colorRamp(c("red", "blue"))
color1(0) ##will  return a matrix of three values - Red  Green  Blue in that order, where it's all red.
color1(1) ##will return a matrix that is all blue. 
color1(0.5) # will return a blend of both red and blue, but no green. 
    color1(seq(0,1, len = 10))   ## generated a sequence, length of 10, of a sequence between all red to all blue. 
    
library(RColorBrewer)
    cols <- brewer.pal(15, "RdYlGn")  ##brewer.pal the only useful argument in this package. colorbrewer.org has the list of available color shchema. 
    pal <-colorRampPalette(cols)  ##rendering the colors into Ramp Pallete
    image(volcano, col= pal(15))  ##asking the colors to be those rendered into pal, and there will be 10 different colors. 

    
x <- rnorm(1000)
y <- rnorm(1000)
plot(x, y, col = rgb(.1,0,.3,.2), pch = 19)  ##RGB looks at red, green, blue as numeric, and last is the alpha of 0.2

```


### Samsung phone- accelerometer example. 
```{r}
list.files("./data/UCI HAR Dataset/")
datapath <-file.path("/Users/payashome/Documents/FMDtRH/R Studio/R Tutorials/R_4_DataScience/data", "UCI HAR Dataset 2")
    files <- list.files(datapath, recursive = TRUE)  ##lists all files in the UCI folder
    files
    
    
library(dplyr)
library(data.table)

## We are going to read in the train, test, features and activities as seperate
#setwd("./data/UCI HAR Dataset 2")
x_train <- read.table(file.path(datapath, "train", "X_train.txt"), header = FALSE)  ##here, file.path saves time by allowing the UCI HAR to be accesses as datapath, in place of pasting the complete filepath. 
  y_train <- read.table(file.path(datapath, "train", "Y_train.txt"), header = FALSE) ## reading in the y train data
      train_sub <- read.table(file.path(datapath, "train", "subject_train.txt"), header = FALSE)
##_________________
x_test <- read.table(file.path(datapath, "test", "X_test.txt"), header = FALSE)  ##reading in data for test, same as above. 
  y_test <- read.table(file.path(datapath, "test", "Y_test.txt"), header = FALSE)
      test_sub <- read.table(file.path(datapath, "test", "subject_test.txt"), header = FALSE)
##_________________
features <- read.table(file.path(datapath, "features.txt"), header = FALSE) ##reading in additional files, no files denoted because they're already in datapath.
actLabel <- read.table(file.path(datapath, "activity_labels.txt"), header = FALSE)

colnames(actLabel) <- c('activityID', 'activityTYPE')
colnames(x_train) = features[ ,2] ##defining the column names as a function of features, which has two columns, so the column name for x_train will come from the list of names in column 2 of features. 
  colnames(y_train) = "activityID"
      y_train <- left_join(y_train, actLabel, by = "activityID")  ##Left joining two data.frames using actLabel as the key, where both column ID's have to match. 
      colnames(train_sub) = "subjectID"

colnames(x_test) = features[ ,2] ##column names are same as in train. 
  colnames(y_test) = "activityID" #descriptive column names for the activity type. 
    y_test <- left_join(y_test, actLabel, by ="activityID")
      colnames(test_sub) = "subjectID" #descriptive column name for the individual, i.e. 1 of 30 participants. 
        ##simply giving column names to the actLabel data.frame
      

combine_train <- cbind(y_train, train_sub, x_train)  ##when str(combine_train), we see activityID as first column (y_train), then the participant ID, then the numerous data columns (x_train)
combine_test <- cbind(y_test, test_sub, x_test)
complete_data <- rbind(combine_train, combine_test) ##row binds the two datasets together. 


names(complete_data) <- gsub("^t", "time", names(complete_data))  ##any string (i.e. column name) beginning with a t changed to time. 
names(complete_data) <- gsub("^f", "frequency", names(complete_data))  ## same for frequency
names(complete_data) <- gsub("Acc", "Accelerometer", names(complete_data))
names(complete_data) <- gsub("Gyro", "Gyroscope", names(complete_data))
names(complete_data) <- gsub("Mag", "Magnitude", names(complete_data))
names(complete_data) <- gsub("BodyBody", "body", names(complete_data)) 

write.table(complete_data, file ="tidydata_samsung", row.name=FALSE)
head(complete_data, n=20)

##___________
par(mfrow = c(1,2), mar = c(5,4,2,1))
complete_data <- transform(complete_data, activityTYPE = factor(activityTYPE))
    sub1 <- subset(complete_data, subjectID == 1)
      plot(sub1[, 4], col= sub1$activityID, ylab = names(sub1)[4])
      plot(sub1[, 5], col = sub1$activityID, ylab = names(sub1)[5])
        op <- par(cex = .5)  ##changes the font size for the legend of the plot. 
              legend("bottomright", legend = unique(sub1$activityTYPE), col = unique(sub1$activityTYPE), pch = 1)

              
par(mfrow = c(1,2), mar = c(5,4,2,1))  ##looking at the max acceleration in x and y for subject 1 
complete_data <- transform(complete_data, activityTYPE = factor(activityTYPE))
    sub1 <- subset(complete_data, subjectID == 1)
      plot(sub1[, 13], col= sub1$activityID, ylab = names(sub1)[13])
      plot(sub1[, 14], col = sub1$activityID, ylab = names(sub1)[14])
        op <- par(cex = .5)  ##changes the font size for the legend of the plot. 
              legend("topright", legend = unique(sub1$activityTYPE), col = unique(sub1$activityTYPE), pch = 19)


```

### EPA Air Quality Example- Fine particulate matter. 
http://goo.gl/soQZHM

```{r}
pm0 <- read.table("./data/PM2-5_2000_2019_annual.txt", header = TRUE, sep= "|", na.strings = "NA", fill = TRUE) ##for a read.table, it expects an element in each column, therefore it is crucial to fill = TRUE in order to fill it in, otherwise it won't be read. 
head(pm0, n= 20)
    
library(dplyr)
    
    pm0 <- pm0[pm0$year %in% c("2000","2019"), ] ##subsetting the data.frame to include two years only. careful, year is denoted as an integer here. strptime not working well for a conversion. 
      #names(pm0) <- names(pm0) %>% make.names() ##renames column names due to abundance of spaces. Fills them in with "_". 
        head(pm0, n=20)
          
        ##taking summary view of sample values. 
        x0 <- pm0$arithmetic_mean
        summary(x0)
        str(x0)
        mean(is.na(x0))
        
        ##subsetting the data into two data.frames, and then boxplotting each on same plot. 
        pm0$year <- as.integer(pm0$year)
            x2000 <- subset(pm0, pm0$year == 2000)
            x2019 <- subset(pm0, pm0$year == 2019)
              boxplot( x2000$arithmetic_mean, x2019$arithmetic_mean, col="blue", pch = 19)

              
              
## ______________ Drilling down into the data_______________##            
        negative <- x2000$arithmetic_mean < 0
        str(negative)
        sum(negative, na.rm = TRUE)
        mean(negative, na.rm = TRUE)
            hist(x2000$arithmetic_mean, xlab = "PM2.5 (micrograms/cubic meter)")  ##getting a view of the data
            site2000 <- unique(subset(x2000, state_code == 36, c(county_code, si_id))) ##returns two column data.frame unique to the state, and showing county and site id's. 
              head(site2000)
              site2000 <- paste(site2000[,1], site2000[,2], sep = ".")  ##pasting together string/number from col 1 to col 2, returning a "number.number" that is the combination of the two. 
        
        negative2 <- x2019$arithmetic_mean < 0
        str(negative)
        sum(negative, na.rm = TRUE)
        mean(negative, na.rm = TRUE)
            hist(x2019$arithmetic_mean, xlab = "PM2.5 (micrograms/cubic meter)")
            site2019 <- unique(subset(x2019, state_code == 36, c(county_code, si_id))) ##returns two column data.frame unique to the state, and showing county and site id's. 
              head(site2019)
              site2019 <- paste(site2019[,1], site2019[,2], sep = ".")
                  both <- intersect(site2019, site2000)  ##here- looks at the county, and site monitors that were present in both 2000 and 2019, and the return shows that there were 7. 
                  str(both)
                  
                  x2000$county_site <- with(x2000, paste(county_code, si_id, sep = ".")) ##creating new columns that are the combo of county and site (from above)
                  x2019$county_site <- with(x2019, paste(county_code, si_id, sep = "."))

                      ct2000 <- subset(x2000, state_code == 36 & county_site %in% both)
                      ct2019 <- subset(x2019, state_code == 36 & county_site %in% both)
                      
                      sapply(split(ct2000, ct2000$county_site), nrow) ## if these were not annual averages, but a collection of dates and n number of observations, the sapply on the split would show us how many observations we had per  station in "both".
                      sapply(split(ct2019, ct2019$county_site), nrow)
```
                      
                      
### EPA Air Quality Example #2- PM2.5

```{r}
key <- readRDS("Source_Classification_Code.rds")
data0 <- readRDS("summarySCC_PM25.rds")
    head(key, n=20)
    head(data0, n=20)
        str(key)
        str(data0)
            summary(data0)
            

library(plotrix)  ##allows us to have a break in the 
with(data0, gap.plot(year, Emissions, gap=c(10,100), col= year, pch = 1, ylim = c(-1, 200), xlab = "Year", ylab = "Emissions", main ="Emissions over time- USA"))
    legend("topright", pch = 1, col = unique(data0$year), legend = unique(data0$year))
    model <- lm(Emissions ~ year, data0) ##creating a liner model between two variables
    abline(model, col = "purple", lty = 2)
        scipen = 5   ##for turning off scientific notation. 
        dev.copy(png, file = "plot1.png")
        dev.off()
        
        
data1 <- subset(data0, fips == "24510")
with(data1, gap.plot(year, Emissions, gap=c(10,100), col= year, pch = 1, ylim = c(-1, 200), xlab = "Year", ylab = "Emissions", main ="Emissions over time- Baltimore City"))
    legend("topright", pch = 1, col = unique(data1$year), legend = unique(data1$year))
    model <- lm(Emissions ~ year, data1) ##creating a liner model between two variables
    abline(model, col = "purple", lty = 2)
        scipen = 5   ##for turning off scientific notation. 
        dev.copy(png, file = "plot2.png")
        dev.off()


        
 ###________________________        
library(ggplot2)   
data2 <- subset(data0, fips == "24510" & type == "POINT")
data3 <- subset(data0, fips == "24510" & type == "NONPOINT")
data4 <- subset(data0, fips == "24510" & type == "ON-ROAD")
data5 <- subset(data0, fips == "24510" & type == "NON-ROAD")

library(RColorBrewer)
    cols <- brewer.pal(6, "RdYlGn")  ##brewer.pal the only useful argument in this package. colorbrewer.org has the list of available color shchema. 
    pal <-colorRampPalette(cols)  ##rendering the colors into Ramp Pallete
    
  
g1 <- ggplot(data2, aes(year, Emissions)) 
p1 <- g1 + geom_point(aes(color = year), size = 4, alpha= 1/2) +   ###assigning a color to a variable requires aes(color = column name), and the alpha is the transparency, size is just that. 
          theme_classic(base_family = "Times", base_size = 9) + ##changes the base font for the entire plot        
            scale_fill_brewer(palette = "RdYlBu") +  ##attempt to use brewer color schemes... doesn't seem to work. 
            geom_smooth(size = 1.5, linetype = 6, color = "orange", method = "lm") + 
            xlab("Year") + ##labels the x axis
            ylab("Emissions") +
            ggtitle("Baltimore City Emissions by 'POINT'") +
            ylim(-1, 30)

g2 <- ggplot(data3, aes(year, Emissions)) 
p2 <- g2 + geom_point(aes(color = year), size = 4, alpha= 1/2) +   ###assigning a color to a variable requires aes(color = column name), and the alpha is the transparency, size is just that. 
          theme_classic(base_family = "Times", base_size = 9) + ##changes the base font for the entire plot        
            scale_fill_brewer(palette = "RdYlBu") +  ##attempt to use brewer color schemes... doesn't seem to work. 
            geom_smooth(size = 1.5, linetype = 6, color = "orange", method = "lm") + 
            xlab("Year") + ##labels the x axis
            ylab("Emissions") +
            ggtitle("Baltimore City Emissions by 'NONPOINT'") +
            ylim(-1, 30)

g3 <- ggplot(data4, aes(year, Emissions)) 
p3 <- g3 + geom_point(aes(color = year), size = 4, alpha= 1/2) +   ###assigning a color to a variable requires aes(color = column name), and the alpha is the transparency, size is just that. 
          theme_classic(base_family = "Times", base_size = 9) + ##changes the base font for the entire plot        
            scale_fill_brewer(palette = "RdYlBu") +  ##attempt to use brewer color schemes... doesn't seem to work. 
            geom_smooth(size = 1.5, linetype = 6, color = "orange", method = "lm") + 
            xlab("Year") + ##labels the x axis
            ylab("Emissions") +
            ggtitle("Baltimore City Emissions by 'ONROAD'") +
            ylim(-1, 30)

g4 <- ggplot(data5, aes(factor(year), Emissions)) 
p4 <- g4 + geom_point(aes(color = year), size = 4, alpha= 1/2) +   ###assigning a color to a variable requires aes(color = column name), and the alpha is the transparency, size is just that. 
          theme_classic(base_family = "Times", base_size = 9) + ##changes the base font for the entire plot        
            scale_fill_brewer(palette = "RdYlBu") +  ##attempt to use brewer color schemes... doesn't seem to work. 
            geom_smooth(size = 1.5, linetype = 6, color = "orange", method = "lm", na.rm= TRUE) + 
            xlab("Year") + ##labels the x axis
            ylab("Emissions") +
            ggtitle("Baltimore City Emissions by 'NONROAD'") +
            ylim(-1, 30)
library(cowplot)

plot_grid(p1, p2, p3, p4, labels= "", ncol=2, nrow=2)
dev.copy(png, file = "plot3.png")
        dev.off()
        
#####___________ Alternative ggplot
data1 <- subset(data0, fips == "24510") ##subset for baltimore
agdata1 <- aggregate(Emissions ~ year, data1, sum)  ##not taht important for this plot.
gg1 <- ggplot(data1, aes(factor(year), Emissions, fill=type)) ##type is a categorical variable/column header. 
gp1 <- gg1 + geom_bar(stat="identity") + ##identity leave the data unchanged. The default is to count
      theme_bw() +
      guides(fill = "none") + ##gets rid of the legend on the right side. 
      facet_grid(.~type, scales = "free", space = "free")+  ##grid is organized by "." or everything, by type(non-road, etc...). Separates 4, 4-olored columns into four grids, each with 4 colors. 
      xlab("Year") + ##labels the x axis
      ylab("Emissions")+
      ggtitle("PM2.5 Emissions, Baltimore City")

gp1
  
  
  
  
        
        library(plotrix)  ##allows us to have a break in the
        library(data.table)
        library(dplyr)
      data6 <- left_join(data0, key, by = "SCC")
data7 <- data6[data6$SCC.Level.Three %like% "coal", ]
with(data7, gap.plot(year, Emissions, gap=c(100,200), col= year, pch = 1, ylim = c(-1, 300), xlab = "Year", ylab = "Emissions", main ="Coal Use over time- USA"))
    legend("topright", pch = 1, col = unique(data6$year), legend = unique(data6$year))
    model <- lm(Emissions ~ year, data7) ##creating a liner model between two variables
    abline(model, col = "purple", lty = 2)
        scipen = 5   ##for turning off scientific notation. 
        dev.copy(png, file = "plot4.png")
        dev.off()
        
        
        
        library(plotrix)  ##allows us to have a break in the
        library(data.table)
        library(dplyr)
        par(mfrow = c(1,1))
data6 <- left_join(data0, key, by = "SCC")
  data8 <- subset(data6, fips == "24510" & SCC.Level.One == "Internal Combustion Engines")
      with(data8, plot(year, Emissions, col= year, pch = 1, cex = 3, xlab = "Year", ylab = "Emissions", main ="Combusion Engine Part. over time- Maryland City"))  ##cex changes the size of the points. 
          legend("topright", pch = 1, col = unique(data8$year), legend = unique(data8$year))
          model <- lm(Emissions ~ year, data8) ##creating a liner model between two variables
              abline(model, col = "purple", lty = 2)
                scipen = 5   ##for turning off scientific notation. 
                dev.copy(png, file = "plot5.png")
                dev.off()
                
          par(mfrow = c(2,1))
                data6 <- left_join(data0, key, by = "SCC")
  data9 <- subset(data6, fips == ("24510") & SCC.Level.One == "Internal Combustion Engines")
      with(data9, plot(year, Emissions, col= fips, pch = 1, xlab = "Year", ylab = "Emissions", main ="Combusion Engine Part. Maryland City"))
          legend("topright", pch = 1, col = unique(data9$fips), legend = unique(data9$fips))
          model <- lm(Emissions ~ year, data9) ##creating a liner model between two variables
          abline(model, col = "purple", lty = 2)
                scipen = 5
 
   data10 <- subset(data6, fips == ("06037") & SCC.Level.One == "Internal Combustion Engines")
      with(data10, plot(year, Emissions, col= fips, pch = 1, xlab = "Year", ylab = "Emissions", main ="Combusion Engine Part. LA County"))
          legend("topright", pch = 1, col = unique(data10$fips), legend = unique(data10$fips))
          model <- lm(Emissions ~ year, data10)
              abline(model, col = "orange", lty = 2)
                scipen = 5   ##for turning off scientific notation. 
                
                dev.copy(png, file = "plot6.png")
                dev.off()
                
```



                      