---
title: "Statistics, Modeling, and Machine Learning"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

## Reproducibility

Steps of data analysis:
  -Define the question
  
  -Define the ideal data set
  
  -Determine what data you can access
  
  -Obtain the data
  
  -Clean the Data
  
  -Exploratory data analysis
  
  -Statistical prediction/Modeling
  
  -Interpret results
  
  -Challenge  results
  
  -Synthesize/Write up results
  
  -Create reproducible code
  
## Probability
  
  P(A u B) = P(A) + P(B)   - The probability of the union between scenario A and B is simply the sum of each.
  Random variable = is the numerical outcome of an experiment (discrete or continuous)
      - e.g. the flip of a coin (discrete random variable)
      - e.g. the roll of a dice (also discrete random variable because it's limited to 1-6)
      - e.g. BMI of a person is a continuous random variable
  
      For random variables, the probabilities must add  up to 1, and components are larger or equal to 0. 
        Probability mass function - for discrete random variables
        Probability density function - for continuous random variables
            in R, probability of some instance occurring within a continuous distribution can be calculated by using pbeta(). Here, p before an argument represents "probability". q before an argument represents "quantile"
            Conditional probability- think lightning strike on a storming vs. sunny day. 
                P(A|B) = P(A u B)/P(B) is the conditional probability. Written as the probability of A given condition B = ....
                
mean = center of a distribution
variance and standard deviations = how spread out the distribution is

```{r }
 ## Beta distribution
pbeta(0.75, 2, 1) ##here 0.75 is 75% probability of some density.
pbeta(c(.4, .5, .6), 2, 1) ##40-60% probability example of the same density.

pnorm(70, mean = 80, sd = 10) ## probability to have value 70 given a set mean and standard dev. Data is normally distributed. 
qnorm(0.95, mean=1100, sd = 75) ## will return a value of a normal distribution given a specificed percentile (95%), and a known mean and st.dev. 

pnorm(16, mean = 15, sd = 1) - pnorm(14, mean = 15, sd = 1) ##probaility that a an output is between two known values, i.e. 14 and 16

ppois(10, lambda = 15) ##poisson distributed data. Probability of seeing 10 or less of an instance, where lamda represents 3 hours,l and 5 people per hour, hence 15. 

```

```{r galton, fig.height=6,fig.width=12, fig.align='center', echo = FALSE, message =FALSE, warning=FALSE}
library(UsingR); data(galton); library(ggplot2)
library(reshape2)
longGalton <- melt(galton, measure.vars = c("child", "parent"))
g <- ggplot(longGalton, aes(x = value)) + geom_histogram(aes(y = ..density..,  fill = variable), binwidth=1, colour = "black") + geom_density(size = 2)
g <- g + facet_grid(. ~ variable)
g
```




```{r }
    g <- ggplot(galton, aes(x = child))
    g <- g + geom_histogram(fill = "salmon", 
      binwidth=1, aes(y = ..density..), colour = "black")
    g <- g + geom_density(size = 2)
    g <- g + geom_vline(xintercept = mean(galton$child), size = 2)
    g
```
 
## Variance = Total sum of squares

```{r}
lambda <- 0.2 
nsim <- 1:1000 # Number of Simulations/rows
n <- 40 

Ematrix <- data.frame(x = sapply(nsim, function(x) {mean(rexp(n, lambda))})) ##creating a matrix of values
head(Ematrix)

Smean <- apply(Ematrix, 2, mean) ## calculating the simulated mean from above matrix.
Smean

Tmean <- 1/lambda  ##calculating the theoretical mean given lambda
Tmean

SSD <- sd(Ematrix$x)  ##The simulated standard deviation
SSD

SVar <- var(Ematrix$x)  ##The simulated variance from the above matrix.
SVar

TSD <- (1/lambda)/sqrt(n)  ##The theoretical standard deviation.
TSD

TVar <- TSD^2   ##The theoretical calculation for the Variance
TVar
```


## t-test

  (X - X*)/(s/sqrt(n)) in other words, sample mean minus the hypothesized/test mean or value divided by standard error of the mean divided by square root of sample size, n. This follows a t-distribution with n-1 degrees of freedom. You can calculate the T-distribution using qt(.95, 15), where .95 is the percentile or an alpha of 0.05, and 15 is n-1 in this example. 

```{r}
library(UsingR); data(father.son)
t.test(father.son$sheight - father.son$fheight)  ##testing whether there are significant differences between the two columns, one for father height and one for son's height. 

library(datasets)
data(ToothGrowth)
head(ToothGrowth)

library(ggplot2)

plot0 <- ggplot(ToothGrowth, aes(supp, len, fill = supp))
plot1 <- plot0 + geom_dotplot(binaxis ="y") +
      facet_grid(.~ dose) 
      plot1

t.test(len ~ supp, data = ToothGrowth) ##testing sig difference between supp types and length

test1 <- subset(ToothGrowth, dose %in% c(.5,1)) ##subsetting doses to make pairwise comparisons
t.test(len ~ dose, data = test1) ##testing sig differences between discrete doses and length

test2 <- subset(ToothGrowth, dose %in% c(1,2)) ##subsetting doses to make pairwise comparisons
t.test(len ~ dose, data = test2)

```

## t-test Confidence Interval
```{r}
## Prompt for below: a sample of 9 men yielded a sample average brain volume of 1,100cc and a standard deviation of 30cc. What is a 95% Student’s T confidence interval for the mean brain vol
mn = 1100
s = 30
n = 9
round(1100 + c(-1,1)*qt(.975, df = 8)*s/sqrt(n))

## diet pill is given to 9 subjects over six weeks. The average difference in weight (follow up - baseline) is -2 pounds. What would the standard deviation of the difference in weight have to be for the upper endpoint of the 95% T confidence interval to touch 0?
n = 9
mn_dif = 2
t = .95
(y_d <- round(mn_dif*sqrt(n) / qt(.975, df = 8), 2))
```


## Two sample t-test
```{r}

##Running a two-sample t-test
data(ChickWeight); library(reshape2)
wideCW <- dcast(ChickWeight, Diet + Chick ~ Time, value.var = "weight") ##dcasting the long-form of the dataset into the wide form by chick, and diet, cast over the time variable, where the represented data is the weight. 
names(wideCW)[-(1:2)] <- paste( "time", names(wideCW)[-(1:2)], sep= "") ##renames columns after the first row, second column, and pastes "time" before the existing column header. 
library(dplyr)
wideCW <- mutate(wideCW, gain = time21-time0)  ##adds a column titled gain, where it subtracts or takes the difference between time 21 and 0, and enters a new column. 

wideCW_14 <- subset(wideCW, Diet %in% c(1,4))
t.test(gain~Diet, paired = FALSE, var.equal = TRUE, data = wideCW_14)
```

## P-Values
```{r}
pt(2.5, 15, lower.tail = FALSE)  ##the probability of getting a T-statistic of 2.5 (first element of the function), with a a df of 15 (i.e. n-1).
```
 
## Least Squares (Regression)

Sum[i=1 to n] of [Yi - (Bo +B1Xi)]^2
Least squares best fit is the sum of i to the n of Yi which is the i'th sample data y-coordinate minus the intercept plus the slope times the i'th sample data x-coordinate, squarred. 
  Bo = intercept of y-axis
  B1 = slope
  Yi = y of sample i
  Xi = x of sample i
        Think of it as subtracting the distance between sample's y-coordinate from the best fit, which is y - (ax + b) which is a linear regression of the best fit, or Yi - Yfit. Then square it.
```{r}
library(UsingR)
library(ggplot2)

freqData <- as.data.frame(table(galton$child, galton$parent))
names(freqData) <- c("child", "parent", "freq")
freqData$child <- as.numeric(as.character(freqData$child))
freqData$parent <- as.numeric(as.character(freqData$parent))
g <- ggplot(filter(freqData, freq > 0), aes(x = parent, y = child))
g <- g  + scale_size(range = c(2, 20), guide = "none" )
g <- g + geom_point(colour="grey50", aes(size = freq+5, show_guide = FALSE))
g <- g + geom_point(aes(colour=freq, size = freq-5))
g <- g + scale_colour_gradient(low = "lightblue", high="white")                    
g



###_____________________Linear Least Squares calculated
y <- galton$child
x <- galton$parent
beta1 <- cor(y, x) *  sd(y) / sd(x)  ##The slope, which is the corelation between y and x times the st.dev of y over the st.dev of x
beta0 <- mean(y) - beta1 * mean(x) ##calculates the y-intercept of the best fit. 
rbind(c(beta0, beta1), coef(lm(y ~ x))) ##lm stands for linear model. coef takes the output and gives us coefficients. REMEMBER - y is the outcome, and x is the predictor. 
##This example gives you the same output, because one is the manual calculation, while the other is done automatically using lm. 

beta1 <- cor(y, x) *  sd(x) / sd(y)  ##if we swapped the predictor 
beta0 <- mean(x) - beta1 * mean(y)
rbind(c(beta0, beta1), coef(lm(x ~ y)))  ##shows us the slope is the same

yc <- y - mean(y)
xc <- x - mean(x)
beta1 <- sum(yc * xc) / sum(xc ^ 2)
c(beta1, coef(lm(y ~ x))[2]) ##returns the same slope

###simplified 
g <- ggplot(filter(freqData, freq > 0), aes(x = parent, y = child))
g <- g  + scale_size(range = c(2, 20), guide = "none" )
g <- g + geom_point(colour="grey50", aes(size = freq+5, show_guide = FALSE))
g <- g + geom_point(aes(colour=freq, size = freq-10))
g <- g + scale_colour_gradient(low = "lightblue", high="white")  
g <- g + geom_smooth(method="lm", col ="orange", formula=y~x) ##for fitting the linear best fit to a plot
g
```

## Linear Regression for Prediction

```{r} 
library(UsingR)
data(diamond)
library(ggplot2)
g = ggplot(diamond, aes(x = carat, y = price))
g = g + xlab("Mass (carats)")
g = g + ylab("Price (SIN $)")
g = g + geom_smooth(method = "lm", colour = "black")
g = g + geom_point(size = 4, colour = "black", alpha=0.5)
g = g + geom_point(size = 2, colour = "blue", alpha=0.2)
g

fit <- lm(price~ carat, data = diamond)  ## prints coefficient Bo and B1, slope and intercept
coef(fit)
summary(fit) ##gives you the complete statistic

fit2 <- lm(price ~ I(carat - mean(carat)), data=diamond) ##how you mean center your predictor, carat. I function lets you provide a calculation in the predictor argument. 
coef(fit2)

fit3 <- lm(price ~ I(carat * 10), data=diamond) ##shortcut where fit3 produces the price change for 1/10 a carat, not a complete carat. 
coef(fit3)


newx <- c(0.16, 0.27, 0.34)  ## an example of a list of carat sizes for diamonds, and we want to predict the price. 
predict(fit, newdata = data.frame(carat = newx))  ##taking original fit data, and calling to the newx concatenated values with the category of carat(carat is what it is referencing). new data is designating new x values.  

```

## Residuals

```{r}
data(diamond)
y <- diamond$price; x <- diamond$carat; n <- length(y)
fit <- lm(y ~ x)
e <- resid(fit)  ##easiest way to calculate residuals
yhat <- predict(fit)
max(abs(e -(y - yhat)))
max(abs(e - (y - coef(fit)[1] - coef(fit)[2] * x)))  ##will return the same as resid()


plot(x, e,  
     xlab = "Mass (carats)", 
     ylab = "Residuals (SIN $)", 
     bg = "lightblue", 
     col = "black", cex = 2, pch = 21,frame = FALSE)
abline(h = 0, lwd = 2)
for (i in 1 : n) 
  lines(c(x[i], x[i]), c(e[i], 0), col = "red" , lwd = 2)





##_______________Example of non-linear data and residuals
x = runif(100, -3, 3); y = x + sin(x) + rnorm(100, sd = .2); 
library(ggplot2)
library(RColorBrewer)
g = ggplot(data.frame(x = x, y = y), aes(x = x, y = y))
g = g + geom_smooth(method = "lm", colour = "black")
g = g + geom_point(size = 7, colour = "black", alpha = 0.4)
g = g + geom_point(size = 5, colour = "red", alpha = 0.4)
g
##_______________
g = ggplot(data.frame(x = x, y = resid(lm(y ~ x))), ##How to plot residuals
           aes(x = x, y = y))
g = g + geom_hline(yintercept = 0, size = 2); 
g = g + geom_point(size = 7, colour = "black", alpha = 0.4)
g = g + geom_point(size = 5, colour = "red", alpha = 0.4)
g = g + xlab("X") + ylab("Residual")
g

##____________ Calculating the residual and plotting for diamond data
diamond$e <- resid(lm(price ~ carat, data = diamond))  ##adding a new column into data.frame for residuals
g = ggplot(diamond, aes(x = carat, y = e))
g = g + xlab("Mass (carats)")
g = g + ylab("Residual price (SIN $)")
g = g + geom_hline(yintercept = 0, size = 2)
g = g + geom_point(size = 7, colour = "black", alpha=0.5)
g = g + geom_point(size = 5, colour = "blue", alpha=0.2)
g
```

## Predictions

```{r}

fit <- lm(mpg~wt, data= mtcars) ## creating a linear model of mtcars dataset, where weight is the predictor
predict(fit, wt=3000) ##using the model above, and predicting the mpg for all cars weighing 3000lbs.

```

## Multivariate Regressions
```{r}

```

## Shapiro Wilks test for Normal distribution 
```{r}
data <- rnorm(100)

#perform Shapiro-Wilk test for normality
shapiro.test(data)  ## where we reject normality when p < 0.05, otherwise we accept that the test data is normally distributed.
```

# Machine Learning

question - input data - features - algorithm - parameters - evaluation

Prediction has accuracy tradeoffs

In-sample error: the error rate you get on the same data set you used to build your predictor. Sometimes called re-substitution error. 
Out of sample error: the error rate you get on a new data set, sometimes called generalization error. 
    Out of sample error is what we should care about
    In sample error is always less than out of sample error
        The reason is over-fitting. 
        True positive vs. false positive, true negative vs. false negative
          Sensitivity - probability that the test positively predicted, and that the prediction was right
          Specificity- probability that the test was negative, and so too was the outcome
  Receiver Operating Characteristic (ROC) - Area under the curve is the measure of goodness of fit (model/prediction to data/outcome). AUC of 0.5 is random guessing. AUC > 0.8 is a good model. 
  Cross-validation. 
    1.) use the training set
    2.) split the training set into a test set and a smaller training set
    3.) Build a model on the training set
    4.) Evaluate the model on the test set
    5.) repeat and average the estimated errors
        k-fold cross-validation is popular technique. 
        
```{r}
library(kernlab)
data(spam)
head(spam)


plot(density(spam$your[spam$type=="nonspam"]),
     col="blue",main="",xlab="Frequency of 'your'")
lines(density(spam$your[spam$type=="spam"]),col="red")
abline(v=0.5,col="black")

##We want to find a value C, which is the cutoff point between spam frequencies, and non-spam frequencies of the word "your". If frequency is above C, predict spam...

prediction <- ifelse(spam$your > 0.5,"spam","nonspam")
table(prediction,spam$type)/length(spam$type)  

```

###Caret for machine learning package

obj Class           Package       Predict Function Syntax
lda                 MASS          Predict(obj) (no options needed)
glm                 stats         predict(obj, type = "response")
bgm                 gbm           predict(obj, type = "response", n.trees)
mda                 mda           predict(obj, typoe ="posterior")
rpart               rpart         predict(obj, type=- "prob")
Weka                RWeka         predict(obj, type= "probability")
LogitBoost          caTools       predict(obj, type= "raw", nIter)

** Different types of predictors have different object class syntax requirements. 


```{r}
library(caret)
library(kernlab) ## to get spam dataset

data(spam)
inTrain <- createDataPartition(y=spam$type, p= 0.75, list = FALSE) ## data is partitioned by spam type, where p is the training proportion. 
training <- spam[inTrain,]  ##subsetting the output of the partition function by spam
testing <- spam[-inTrain,]  ##subsetting the output of the partition function by not spam
dim(training)  ##shows the dimensions of the dataframe

set.seed(11111)
modelFit <- train(type ~., data= training, method = "glm")  ##training a model, specifically a glm model, where we are comparing column type (spam vs. not spam) with everything ".", and using the subsetted data for training. 
modelFit

modelFit$finalModel  ##will return the fitted values for all of the other columns of data. The higher the fit, the more likely it's spam. 

predictions <- predict(modelFit, newdata = testing)  ##how we can predict on new samples, here using the subset testing data.frame. 
predictions
confusionMatrix(predictions, testing$type)  ##using the confusingMatrix argument to determine how well the predictions are working, and the focus is on predicting type (spam vs. not spam)

```

### Data Slicing with caret
```{r}
library(caret)
library(kernlab) ## to get spam dataset

data(spam)
inTrain <- createDataPartition(y=spam$type, p= 0.75, list = FALSE) ## data is partitioned by spam type, where p is the training proportion. 
training <- spam[inTrain,]  ##subsetting the output of the partition function by spam
testing <- spam[-inTrain,]  ##subsetting the output of the partition function by not spam
dim(training)

set.seed(11111)
folds <- createFolds(y= spam$type, k=10, list= TRUE, returnTrain=TRUE)  ##k is the number of folds, smap$type is the column being references. You can return either the Training set, or the test set. 
sapply(folds, length)

```

## Plotting Predictors + Hmisc for cutting data.frames
```{r}
library(ISLR)
library(ggplot2)
library(caret)

data(wage)
summary(Wage)
head(Wage, n=20)

inTrain <- createDataPartition(y=Wage$wage, p=0.7, list=FALSE)
training <- Wage[inTrain,]
testing <- Wage[-inTrain,]
dim(training); dim(testing)  ##returns the number of rows and columns

featurePlot(x=training[, c("age", "education", "jobclass")], ##a plotting function that comes with caret, and we're taking a subset of the data, three columns in this instance.
            y= training$wage,  ##this is your output of interest
            plot="pairs")

qplot(age, wage, data=training) ##here we see two distinct groups
qplot(age, wage, colour= jobclass, data= training)

qp <- qplot(age, wage, color= education, data=training)
qp + geom_smooth(method = 'lm', formula= y~x)

library(Hmisc)  ##good library for cutting data.frames into pieces. 
cutWage <- cut2(training$wage, g=3) ##cut2 argument performs the cut, by $wage column, into g=3 pieces. 
table(cutWage)

p1 <- qplot(cutWage, age, data=training, fill=cutWage, geom=c("boxplot"))
p1          

t1 <- table(cutWage, training$jobclass)  ##great way to make quick tables
t1

prop.table(t1,1)  ##proportion table, where the 1 represents proportion in each row. a 2 would be each column. 

qplot(wage, color=education, data= training, geom="density")
```

### Pre-processing predictor variables
```{r}
library(caret)
library(kernlab) ## to get spam dataset
library(RANN)

data(spam)
inTrain <- createDataPartition(y=spam$type, p= 0.75, list = FALSE) ## data is partitioned by spam type, where p is the training proportion. 
training <- spam[inTrain,]  ##subsetting the output of the partition function by spam
testing <- spam[-inTrain,]  ##subsetting the output of the partition function by not spam
dim(training)
hist(training$capitalAve, main="", xlab="ave.capital run length")
mean(training$capitalAve)
sd(training$capitalAve) ## what we see here is that the standard deviation is significanly higher than the mean, which indicates that we need some pre-processing. 

trainCapAve <- training$capitalAve
trainCapAveS <- (trainCapAve - mean(trainCapAve))/sd(trainCapAve) ## A way of standardizing the data
mean(trainCapAveS)
sd(trainCapAveS)

##If we want to then standardize the test set, we must use the mean from the training set, and the st.dev from the training set. 
testCapAve <- testing$capitalAve
testCapAveS <- (testCapAve - mean(trainCapAve))/sd(trainCapAve) 
mean(testCapAveS)
sd(testCapAveS)

##Alternative is passing preProcess into the train() argument
modelFit <- train(type~., data= training, preProcess= c("center", "scale"), method="glm")  ##This processes for all of the predictor variables. 
modelFit


## Prediction models tend to fail with NA. Well use K nearest imputation statistical function. 
# Make some values NA
training$capAve <- training$capitalAve
selectNA <- rbinom(dim(training)[1],size=1,prob=0.05)==1  ## we are adding NA's into the dataset here. 
training$capAve[selectNA] <- NA
# Impute and standardize
preObj <- preProcess(training[,-58],method="knnImpute")  ##removing the 58th column here
capAve <- predict(preObj,training[,-58])$capAve
# Standardize true values
capAveTruth <- training$capitalAve
capAveTruth <- (capAveTruth-mean(capAveTruth))/sd(capAveTruth)

```

### Covariate (predictor) creation
```{r}
 ##think about creating or calculating variables/features to summarize the sample data, in this instance an entire e-mail if we use the spam dataset. 

library(ISLR); library(caret); data(Wage);
inTrain <- createDataPartition(y=Wage$wage,
                              p=0.7, list=FALSE)
training <- Wage[inTrain,]
testing <- Wage[-inTrain,]

##turning qualitative variables into quantitative ones
table(training$jobclass)
dummies <- dummyVars(wage ~ jobclass,data=training)  ##using the dummyVars argument to turn categorical info, such as "industrial" or "information" into integers. Outcome is wage, and jobclass is the predictor.
head(predict(dummies,newdata=training)) 

##__________________
##A way to throw out less-meaningful predictors, i.e. near-zero variations in the data. 
nsv <- nearZeroVar(training,saveMetrics=TRUE)
nsv  ##in this example we can throw out sex and region


```

### Predicting with Regression
```{r}
##Using the old faithful dataset
library(caret);data(faithful); set.seed(333)
inTrain <- createDataPartition(y=faithful$waiting,
                              p=0.5, list=FALSE)
trainFaith <- faithful[inTrain,]; testFaith <- faithful[-inTrain,]
head(trainFaith)

plot(trainFaith$waiting,trainFaith$eruptions,pch=19,col="blue",xlab="Waiting",ylab="Duration") ##plotting the training data. 
lm1 <- lm(eruptions ~ waiting, data=trainFaith)
summary(lm1)
##output is y(eruption duration) = 0.073 * (waiting time) - 1.792

## To predict a new value, we can automate this by:
coef(lm1)[1] + coef(lm1)[2]*80  ## coef(lm1)[1] returns the intercept, and [2] returns the slope. We are predicting with a wait time of 80

newdata <- data.frame(waiting=80)
predict(lm1,newdata)  ## a shortcut so we don't have to continuously calculate the coeficients.

##We can not use the predictions from the training set on the TEST set
pred1 <- predict(lm1,newdata=testFaith,interval="prediction")
ord <- order(testFaith$waiting)
plot(testFaith$waiting,testFaith$eruptions,pch=19,col="blue")

##________________
##We can use CARET to do the same, and much faster
modFit <- train(eruptions ~ waiting,data=trainFaith,method="lm")  ##eruptions is outcome, waiting is predictor. model built on the training dataset, and method is linear modeling. 
summary(modFit$finalModel)  ##How we get final model output

```

### Predicting Multiple Covariate Regression
```{r}
library(ISLR); library(ggplot2); library(caret);
data(Wage); Wage <- subset(Wage,select=-c(logwage))  ##here we are subsetting and removing the variable we are trying to predict
summary(Wage)



inTrain <- createDataPartition(y=Wage$wage,##subsetting into test and train
                              p=0.7, list=FALSE)
training <- Wage[inTrain,]
testing <- Wage[-inTrain,]
dim(training); dim(testing)


featurePlot(x=training[,c("age","education","jobclass")],
            y = training$wage,
            plot="pairs")

##________________

modFit<- train(wage ~ age + jobclass + education,
               method = "lm",data=training)
finMod <- modFit$finalModel
print(modFit)


plot(finMod,1,pch=19,cex=0.5,col="#00000010")  ##residuals vs. fitted. We want to see a straight line
qplot(finMod$fitted,finMod$residuals,colour=race,data=training)  ##also fitted vs. residuals. Trying to identify different trends here. 



```

### Predicting with Trees
Better performance in non-linear settings
Can cause over-fitting, so be careful 
Considered non-linear models, so they use interactions between variables

1.) start with all variables in one group
2.) Find the variable/split that best separates the outcomes
3.) Divide the data into two groups "leaves" on that node
4.) Within each split, find the best variable/split that separates the outcomes
5.) Continue until the groups are too small or sufficiently "pure"
```{r}
data(iris); library(ggplot2)
names(iris)
table(iris$Species)


inTrain <- createDataPartition(y=iris$Species,
                              p=0.7, list=FALSE)
training <- iris[inTrain,]
testing <- iris[-inTrain,]
dim(training); dim(testing)

qplot(Petal.Width,Sepal.Width,colour=Species,data=training)

library(caret)
modFit <- train(Species ~ .,method="rpart",data=training) ##train function fits the model here. looking at "." or everything as predictors, and species is the outcome, method = r's package for regression/classification. 
print(modFit$finalModel)

plot(modFit$finalModel, uniform=TRUE, 
      main="Classification Tree")
text(modFit$finalModel, use.n=TRUE, all=TRUE, cex=.8)  ##plotting the dendogram tree

library(rattle)
fancyRpartPlot(modFit$finalModel)  ##makes a fancier version of the same dendogram
```

### Bootstrap-aggregating "Bagging"
Re-sample the data, and recalculate predictions, then average or majority vote.
Useful for non-linear functions
When using bagging in caret, consider using "bagEarth", "treebag", "bagFDA" inside of the train function. 

Below is the syntax/code, but ozone dataset not available
predictors = data.frame(ozone=ozone$ozone)
temperature = ozone$temperature
treebag <- bag(predictors, temperature, B = 10,
                bagControl = bagControl(fit = ctreeBag$fit,
                                        predict = ctreeBag$pred,
                                        aggregate = ctreeBag$aggregate))


### Random Forests
1.) Bootstrap samples
2.) at each split, bootstrap the variables again
3.) Grow multiple trees and vote
Cons- prone to overfitting, but it is very accurate
```{r}
data(iris); library(ggplot2); library(randomForest); library(caret)
inTrain <- createDataPartition(y=iris$Species,
                              p=0.7, list=FALSE)
training <- iris[inTrain,]
testing <- iris[-inTrain,]


modFit <- train(Species~ .,data=training,method="rf",prox=TRUE)  ##using train, here method is "rf", which stands for random forest. Outcome is species, and predictors are every other variable. 
modFit

getTree(modFit$finalModel,k=2) ##to view the individual trees, you can use this function, which is looking at model 2


irisP <- classCenter(training[,c(3,4)], training$Species, modFit$finalModel$prox)  ##This centers your class predictions, which will be used in the plot below
irisP <- as.data.frame(irisP); irisP$Species <- rownames(irisP)
p <- qplot(Petal.Width, Petal.Length, col=Species,data=training)
p + geom_point(aes(x=Petal.Width,y=Petal.Length,col=Species),size=5,shape=4,data=irisP)


pred <- predict(modFit,testing); testing$predRight <- pred==testing$Species
table(pred,testing$Species)

qplot(Petal.Width,Petal.Length,colour=predRight,data=testing,main="newdata Predictions")
```

### Boosting
1.) Takes a lot of weak predictors
2.) Weight them and add them up
3.) Get a stronger predictor
```{r}
library(ISLR); data(Wage); library(ggplot2); library(caret);
Wage <- subset(Wage,select=-c(logwage))  ##removes the variable were interested in
inTrain <- createDataPartition(y=Wage$wage,
                              p=0.7, list=FALSE)
training <- Wage[inTrain,]; testing <- Wage[-inTrain,]

modFit <- train(wage ~ ., method="gbm",data=training,verbose=FALSE)  ## "gbm" is boosting with trees using the train function from caret. 
print(modFit)  

qplot(predict(modFit,testing),wage,data=testing)
```

## Practical Example 1
1. Subset the data to a training set and testing set based on the Case variable in the data set.  
2. Set the seed to 125 and fit a caret model with the rpart method using all predictor variables and default caret settings.  
3. In the final model what would be the final model prediction for cases with the following variable values:  
  a. TotalIntench2 = 23,000; FiberWidthCh1 = 10; PerimStatusCh1 = 2.  
  b. TotalIntench2 = 50,000; FiberWidthCh1 = 10; VarIntenCh4 = 100.  
  c. TotalIntench2 = 57,000; FiberWidthCh1 = 8; VarIntenCh4 = 100.  
  d. FiberWidthCh1 = 8; VarIntenCh4 = 100;  PerimStatusCh1 = 2. 
```{r}
library(rattle)
library(AppliedPredictiveModeling) 
data(segmentationOriginal)
seg1 <- segmentationOriginal
head(seg1, n=20)

set.seed(125)

# split train and test
library(dplyr)
training <- dplyr::filter(seg1, Case =="Train")
testing <- dplyr::filter(seg1, Case =="Test")

# fit CART model with rpart
fit <- train(Class ~ ., method="rpart", data=training)

fit$finalModel  


fancyRpartPlot(fit$finalModel)  ##fancy version, but requires a method of rpart to plot it.
```
### Practical example 2
```{r}
# Set seed for reproducibility
set.seed(1111)

## Data processing- In this section the data is downloaded and manipulated. Some basic transformations and cleanup will be performed, so that `NA` values are omitted. 

training   <-read.csv("./data/pml-training.csv", na.strings=c("NA","#DIV/0!", "")) ##Cleaning
testing <-read.csv("./data/pml-testing.csv", na.strings=c("NA", "#DIV/0!", ""))
training<-training[,colSums(is.na(training)) == 0]
testing <-testing[,colSums(is.na(testing)) == 0]

training   <-training[,-c(1:7)]  ##Subset for just the columns of interest. 
testing <-testing[,-c(1:7)]


# Cross-validation: In this section cross-validation will be performed by splitting the training data in training (75%) and testing (25%) data.
library(caret)
subSamples <- createDataPartition(y=training$classe, p=0.75, list=FALSE)
subTraining <- training[subSamples, ] 
subTesting <- training[-subSamples, ]


library(ggplot2) ##some exploration
p1 <- ggplot(subTraining, aes(classe))
plot <- p1 + geom_boxplot()
plot

## Prediction models: In this section a decision tree and random forest will be used on the data.
library(rattle)
# Fit model
modFitDT <- train(classe ~ ., method="rpart", data=subTraining)
# Perform prediction
predictDT <- predict(modFitDT, subTesting, type = "prob")
# Plot result

fancyRpartPlot(modFitDT$finalModel)


##This Random forest takes 1.5hrs to run, be careful!!!!!!!!!
# Fit model
library(caret)
library(randomForest)
modFit <- train(classe~ .,data=subTraining,method="rf")  
modFit

gTree <- getTree(modFit$finalModel,k=3) ##to view the individual trees, you can use this function, which is looking at model 3
head(gTree, n=30)

predictRF <- predict(modFit, subTesting, type = "prob")

# Perform prediction
predicts <- predict(modFit, testing, type="prob")
predicts
```



### Forecasting
Traditionally independent variable over TIME
```{r}

```

